---
title: "COVID-19 Analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

Based on COVID-19 case and death data from the [New York Times COVID-19 data repository](https://github.com/nytimes/covid-19-data) and population data from [US Census Bureau](https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/totals/).

```{r echo = FALSE}
covid_st <- read.csv("../nytimes/covid-19-data/us-states.csv")
covid_st$date <- as.Date(covid_st$date)

covid_co <- read.csv("../nytimes/covid-19-data/us-counties.csv")
covid_co$date <- as.Date(covid_co$date)

pop_co <- read.csv("../census/co-est2019-alldata.csv")[, c("STATE", "COUNTY", "STNAME", "CTYNAME", "POPESTIMATE2019")]
pop_st <- read.csv("../census/SCPRC-EST2019-18+POP-RES.csv")[, c("STATE", "NAME", "POPESTIMATE2019")]

# https://gist.github.com/dantonnoriega/bf1acd2290e15b91e6710b6fd3be0a53

postal <- read.csv("../dantonnoriega/us-state-ansi-fips.csv")

# merge state data

m_st <- merge(x=covid_st, y=pop_st, by.x="fips", by.y="STATE", all.x=FALSE, all.y=FALSE)
m_st <- merge(x=m_st, y=postal, by.x="fips", by.y="st")

# add data for last week

m_st$w1 <- m_st$date - 7
m_st <- merge(x=m_st, y=m_st[, c("date", "fips", "cases", "deaths")], by.x=c("fips", "w1"), by.y=c("fips", "date"), all.x=TRUE, suffixes=c("", ".w1"))

# add data for two weeks ago

m_st$w2 <- m_st$date - 14
m_st <- merge(x=m_st, y=m_st[, c("date", "fips", "cases", "deaths")], by.x=c("fips", "w2"), by.y=c("fips", "date"), all.x=TRUE, suffixes=c("", ".w2"))

# calculate per capita rates

m_st$cases_permil <- m_st$cases / m_st$POPESTIMATE2019 * 1000000.
m_st$deaths_permil <- m_st$deaths / m_st$POPESTIMATE2019 * 1000000.

m_st$new_cases <- m_st$cases - m_st$cases.w1
m_st$new_deaths <- m_st$deaths - m_st$deaths.w1
m_st$new_cases_permil <- m_st$new_cases / m_st$POPESTIMATE2019 * 1000000.
m_st$new_deaths_permil <- m_st$new_deaths / m_st$POPESTIMATE2019 * 1000000.

m_st$new_case_rate <- m_st$cases - 2 * m_st$cases.w1 + m_st$cases.w2
m_st$new_death_rate <- m_st$deaths - 2 * m_st$deaths.w1 + m_st$deaths.w2
m_st$new_case_rate_permil <- m_st$new_case_rate / m_st$POPESTIMATE2019 * 1000000.
m_st$new_death_rate_permil <- m_st$new_death_rate / m_st$POPESTIMATE2019 * 1000000.

# calculate ranks for each date

for (date in min(m_st$date):max(m_st$date)) {
  m_st[m_st$date == date, "c_rank"] <- rank(m_st[m_st$date == date, "cases_permil"])
  m_st[m_st$date == date, "d_rank"] <- rank(m_st[m_st$date == date, "deaths_permil"])
  m_st[m_st$date == date, "new_c_rank"] <- rank(m_st[m_st$date == date, "new_cases_permil"])
  m_st[m_st$date == date, "new_d_rank"] <- rank(m_st[m_st$date == date, "new_deaths_permil"])
  m_st[m_st$date == date, "new_c_rate_rank"] <- rank(m_st[m_st$date == date, "new_case_rate_permil"])
  m_st[m_st$date == date, "new_d_rate_rank"] <- rank(m_st[m_st$date == date, "new_death_rate_permil"])
}

# merge county data

pop_co$FIPS = pop_co$STATE * 1000 + pop_co$COUNTY
m_co <- merge(x=covid_co, y=pop_co[, c("FIPS", "POPESTIMATE2019") ], by.x="fips", by.y="FIPS",
              all.x=FALSE, all.y=FALSE)
m_co$cases_permil <- m_co$cases / m_co$POPESTIMATE2019 * 1000000.
m_co$deaths_permil <- m_co$deaths / m_co$POPESTIMATE2019 * 1000000.
```

States of interest:

```{r echo = FALSE}
states <- c("California", "Mississippi", "Illinois", "District of Columbia", "New York", "Massachusetts", "Texas", "North Carolina", "Louisiana", "Alabama")
states
```

## Cumulative Analysis

State data as of:

```{r echo = FALSE}
last_st <- max(m_st$date)
last_st
t_st <- m_st[m_st$date == last_st, ]
```

Total cases:

```{r echo = FALSE}
sum(t_st$cases)
```

Total deaths:

```{r echo = FALSE}
sum(t_st$deaths)
```

State case rank:

```{r echo = FALSE}
c_order_st <- order(t_st$c_rank)
t_st[c_order_st, c("state", "cases", "cases_permil", "c_rank")]
```

Case rank over time since March 1:

```{r echo = FALSE, fig.height=8}
m_st_sel <- m_st[is.element(m_st$state, states) & m_st$date >= "2020-03-01", ]
xrange <- range(m_st_sel$date)
xrange[2] <- xrange[2] + 7 # pad to leave space on right for labels
yrange <- range(m_st_sel$c_rank)
plot(xrange, yrange, type="n", xlab="Date",
   ylab="Per Capita Case Rank", bty="n")
pallette <- rainbow(length(states))
names(pallette) <- states
for (state in states) {
  st_df <- m_st[m_st$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  lines(st_df$date, smooth(st_df$c_rank, kind="3RS3R", endrule="copy"), col=pallette[state])
  last <- st_df[nrow(st_df), ]
  text(last$date, y=last$c_rank, last$stusps, pos=4, offset=0.1, cex=0.8, col=pallette[state])
} 

```

State death rank:

```{r echo = FALSE}
d_order_st <- order(t_st$d_rank)
t_st[d_order_st, c("state", "deaths", "deaths_permil", "d_rank")]
```

Death rank over time since March 1:

```{r echo = FALSE, fig.height=8}
m_st_sel <- m_st[is.element(m_st$state, states) & m_st$date >= "2020-03-01", ]
xrange <- range(m_st_sel$date)
xrange[2] <- xrange[2] + 7 # pad to leave space on right for labels
yrange <- range(m_st_sel$d_rank)
plot(xrange, yrange, type="n", xlab="Date",
   ylab="Per Capita Death Rank", bty="n")
pallette <- rainbow(length(states))
names(pallette) <- states
for (state in states) {
  st_df <- m_st[m_st$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  lines(st_df$date, smooth(st_df$d_rank, kind="3RS3R", endrule="copy"), col=pallette[state])
  last <- st_df[nrow(st_df), ]
  text(last$date, y=last$d_rank, last$stusps, pos=4, offset=0.1, cex=0.8, col=pallette[state])
} 

```

County data as of:

```{r echo = FALSE}
last_co <- max(m_co$date)
t_co <- m_co[m_co$date == last_co, ]
last_co
```
Total counties:

```{r echo = FALSE}
nrow(t_co)
```

Counties from selected states with death rank in bottom 100:

```{r echo = FALSE}
t_co$d_rank <- rank(t_co$deaths_permil)
d_order_co <- order(t_co$d_rank)

t_co_100 <- t_co[tail(d_order_co, 100), c("state", "county", "deaths", "deaths_permil", "d_rank")]
t_co_100[is.element(t_co_100$state, states), ]
```

Selected state contributions to worst 100 counties:

```{r echo = FALSE}
st_co_table <- table(state=t_co[tail(d_order_co, 100), "state"])
st_co_table_df <- as.data.frame(st_co_table, responseName="counties")
st_co_table_df <- st_co_table_df[is.element(st_co_table_df$state, states), ]
st_co_table_df[order(st_co_table_df$counties), ]
```

## Latest Week Analysis

State new case rank:

```{r echo = FALSE}
new_c_order_st <- order(t_st$new_c_rank)
t_st[new_c_order_st, c("state", "new_cases", "new_cases_permil", "new_c_rank")]
```

New case rank over time since March 1:

```{r echo = FALSE, fig.height=8}
m_st_sel <- m_st[is.element(m_st$state, states) & m_st$date >= "2020-03-01", ]
xrange <- range(m_st_sel$date)
xrange[2] <- xrange[2] + 7 # pad to leave space on right for labels
yrange <- range(m_st_sel$new_c_rank)
plot(xrange, yrange, type="n", xlab="Date",
   ylab="Per Capita New Case Rank", bty="n")
pallette <- rainbow(length(states))
names(pallette) <- states
for (state in states) {
  st_df <- m_st[m_st$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  lines(st_df$date, smooth(st_df$new_c_rank, kind="3RS3R", endrule="copy"), col=pallette[state])
  last <- st_df[nrow(st_df), ]
  text(last$date, y=last$new_c_rank, last$stusps, pos=4, offset=0.1, cex=0.8, col=pallette[state])
} 

```

State death rank:

```{r echo = FALSE}
new_d_order_st <- order(t_st$new_d_rank)
t_st[new_d_order_st, c("state", "new_deaths", "new_deaths_permil", "new_d_rank")]
```

New death rank over time since March 1:

```{r echo = FALSE, fig.height=8}
m_st_sel <- m_st[is.element(m_st$state, states) & m_st$date >= "2020-03-01", ]
xrange <- range(m_st_sel$date)
xrange[2] <- xrange[2] + 7 # pad to leave space on right for labels
yrange <- range(m_st_sel$new_d_rank)
plot(xrange, yrange, type="n", xlab="Date",
   ylab="Per Capita New Death Rank", bty="n")
pallette <- rainbow(length(states))
names(pallette) <- states
for (state in states) {
  st_df <- m_st[m_st$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  lines(st_df$date, smooth(st_df$new_d_rank, kind="3RS3R", endrule="copy"), col=pallette[state])
  last <- st_df[nrow(st_df), ]
  text(last$date, y=last$new_d_rank, last$stusps, pos=4, offset=0.1, cex=0.8, col=pallette[state])
} 

```

## All States

```{r echo=FALSE}
states <- unique(m_st$state)
cols <- 4
rows <- floor((length(states) + 1) / cols)
```

### Cumulative Cases Per Million

```{r echo=FALSE, fig.height=20}
par(mfrow=c(rows, cols))
par(mar=c(2, 2, 2, 1))
xrange <- range(m_st$date)
yrange <- range(m_st$cases_permil)
for (state in t_st$state[c_order_st]) {
  st_df <- m_st[m_st$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  main <- sprintf("%.0f %s", st_df$c_rank[nrow(st_df)], st_df$stusps[1])
  plot(xrange, yrange, type="n", main=main, xlab="", ylab="")
  lines(st_df$date, st_df$cases_permil)
}
```

### Cumulative Deaths Per Million

```{r echo=FALSE, fig.height=20}
par(mfrow=c(rows, cols))
par(mar=c(2, 2, 2, 1))
xrange <- range(m_st$date)
yrange <- range(m_st$deaths_permil)
for (state in t_st$state[d_order_st]) {
  st_df <- m_st[m_st$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  main <- sprintf("%.0f %s", st_df$d_rank[nrow(st_df)], st_df$stusps[1])
  plot(xrange, yrange, type="n", main=main, xlab="", ylab="")
  lines(st_df$date, st_df$deaths_permil)
}
```

### New Cases Per Million

```{r echo=FALSE, fig.height=20}
par(mfrow=c(rows, cols))
par(mar=c(2, 2, 2, 1))
df <- m_st[!is.na(m_st$new_cases_permil),]
xrange <- range(df$date)
yrange <- range(df$new_cases_permil)
for (state in t_st$state[new_c_order_st]) {
  st_df <- df[df$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  main <- sprintf("%.0f %s", st_df$new_c_rank[nrow(st_df)], st_df$stusps[1])
  plot(xrange, yrange, type="n", main=main, xlab="", ylab="")
  lines(st_df$date, st_df$new_cases_permil)
}
```

### New Deaths Per Million

```{r echo=FALSE, fig.height=20}
par(mfrow=c(rows, cols))
par(mar=c(2, 2, 2, 1))
df <- m_st[!is.na(m_st$new_deaths_permil),]
xrange <- range(df$date)
yrange <- range(df$new_deaths_permil)
for (state in t_st$state[new_d_order_st]) {
  st_df <- df[df$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  main <- sprintf("%.0f %s", st_df$new_d_rank[nrow(st_df)], st_df$stusps[1])
  plot(xrange, yrange, type="n", main=main, xlab="", ylab="")
  lines(st_df$date, st_df$new_deaths_permil)
}
```

### New Case Rate Per Million

```{r echo=FALSE, fig.height=20}
new_c_rate_order_st <- order(t_st$new_c_rate_rank)
par(mfrow=c(rows, cols))
par(mar=c(2, 2, 2, 1))
df <- m_st[!is.na(m_st$new_case_rate_permil),]
xrange <- range(df$date)
yrange <- range(df$new_case_rate_permil)
for (state in t_st$state[new_c_rate_order_st]) {
  st_df <- df[df$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  main <- sprintf("%.0f %s", st_df$new_c_rate_rank[nrow(st_df)], st_df$stusps[1])
  plot(xrange, yrange, type="n", main=main, xlab="", ylab="")
  lines(st_df$date, st_df$new_case_rate_permil)
}
```

### New Death Rate Per Million

```{r echo=FALSE, fig.height=20}
new_d_rate_order_st <- order(t_st$new_d_rate_rank)
par(mfrow=c(rows, cols))
par(mar=c(2, 2, 2, 1))
df <- m_st[!is.na(m_st$new_death_rate_permil),]
xrange <- range(df$date)
yrange <- range(df$new_death_rate_permil)
for (state in t_st$state[new_d_rate_order_st]) {
  st_df <- df[df$state == state, ]
  st_df <- st_df[order(st_df$date), ]
  main <- sprintf("%.0f %s", st_df$new_d_rate_rank[nrow(st_df)], st_df$stusps[1])
  plot(xrange, yrange, type="n", main=main, xlab="", ylab="")
  lines(st_df$date, st_df$new_death_rate_permil)
}
```